Extracellular potentials generated by axonal projections are shaped by patterns of bifurcations and terminations

### 2016-02-02 doc/manuscript.tex
add more explanation to results and call figures in order.

These properties are first investigated by numerical model based on a
forward simulation of the extracellular field potential (Holt and Koch,
1999; Gold et al., 2006). This first model includes a large-scale
multicompartment model of the axon population. We then show a concrete
example of the effects in experimental data using as a set of multi-site
in-vivo electrophysiological recordings from the barn owl brain stem.
Finally, we show the mechanisms underlying the observed effects by means
of a second, analytically tractable, model of a generic axon bundle.
Results ==============

\emph{Axonal projections generate a dipole-like field potential.} To
understand how axons can generate a dipolar EFP, we simulated a parallel
fiber bundle that at first runs at constant diameter and without
bifurcations, then reaches a bifurcation zone with an increasing number
of fibers, and finally reaches a termination zone, in which the number
of fibers decreases to zero. (Figure 1A and 1D). The bundle fibers were
stimulated with a background spontaneous rate, upon which a pulse of
increased activity was added. To characterize the spatiotemporal
dynamics of the evoked EFP, the potential was calculated for several
locations along the axon trunk.

The spatial behaviour of the low-frequency component of the EFP differed
strongly from that of the high-frequency component, so we analyzed them
separately. Low-frequency responses are shown in the top half of Figure
1, and high-frequency responses are shown in the lower half.

The distribution of maximum amplitudes, shown by the contour lines in
Figure 1A, showed the double-lobed shape typical of a dipole, and
exceeds the reach of the high-frequency maximum responses (Fig. 1D). The
low-frequency waveforms (Fig 1B) show a biphasic deflection that is
elicited by the population firing rate pulse. This deflection was
visible at all recoring locations. It reversed polarity in the middle of
the bifurcation zone of the bundle (red box in Figure 1B).

The same distribution of maximum amplitudes for the high-frequency
component, shown in Figure 1D, did not show the same lobed shape.
Instead it decayed much faster with distance in both axial and radial
direction. This was also reflected in the waveforms (Fig 1E). There is
an increase in amplitude due to the firing rate pulse at locations close
to the nerve bundle, but at locations further away from the trunk it is
no longer distinguishable from the background activity.

@@

\emph{Frequency components are related to different anatomical
features.} To better understand the relationship between axonal anatomy
and this spatial structure of the EFP, we plotted the change in density
of the bundle, meaning the number of terminations subtracted from the
number of bifurcations at a given

The low-frequency component shows a distinct relationship to the
anatomy: it appears to be governed by the local density of bifurcations
and terminations. Along the nerve trunk the fiber density is constant,
and the low frequency component is unchanged. As the axon bundle reaches
its projection zone, the number of bifurcations increases, and the low
frequency component increases in amplitude. In the middle of the
projection zone, the number of bifurcations and terminations cancels
out. Correspondingly, the low-frequency component amplitude reaches a
local minimum. As the end of the projection zone approaches, the
terminations outweigh the bifurcations, and the amplitude increases, but
the polarity is reversed (red box in Figure 1B). As the axon bundle
ends, there are no longer any bifurcations or terminations, and the
amplitude decays.

The high-frequency component changes its amplitude in the longitudinal
direction in accordance to the local fiber density. It has an
approximately constant amplitude along the nerve trunk and then
increases in amplitude as the fiber density is increased by
bifurcations. As the fibers terminate and fiber density decreases, so
does the amplitude of the high-frequency component.

These relationships suggest that the particular structure of the axon
bundle gives rise to the dipolar field potential. Because dipolar fields
are associated with a longer spatial reach that the quarupolar fields
typically associated with axonal spikes, we decided to investigate the
effect that axon structure has on the spatial reach of the frequency
components. To do this we performed a further simulation of an axon
bundle, this time omitting the bifurcations, and letting the fibers
terminate directly. The resulting distance decay profiles are shown in
Figure 2. In axial direction, the low-frequency component has a further
reach than the high frequency component in the bifurcating case (Figure
2A), but not in the non-bifurcating case (Figure 2C). In the radial
direction, the situation is different: the bifurcating (Figure 2B) and
non-bifurcating (Figure 2D) case both show a faster decay of the
high-frequency component than of the low-frequency component. To assure
that the effect was due to the bifurcations themselves and not due to
the fanned out structure of the bifurcating bundle seen in Figures 2A
and 2D, we here changed the layout of both bundles to have constant
widths.

>>>>

These properties are first investigated by numerical model based on a
forward simulation of the extracellular field potential (Holt and Koch,
1999; Gold et al., 2006). This first model includes a large-scale
multicompartment model of the axon population. We then show a concrete
example of the effects in experimental data using as a set of multi-site
in-vivo electrophysiological recordings from the barn owl brain stem.
Finally, we show the mechanisms underlying the observed effects by means
of a second, analytically tractable, model of a generic axon bundle.

\section{Results}\label{results}

\emph{Axonal projections generate a dipole-like field potential.} To
understand how axons can generate a dipolar EFP, we simulated a parallel
fiber bundle that at first runs at constant diameter and without
bifurcations, then reaches a bifurcation zone with an increasing number
of fibers, and finally reaches a termination zone, in which the number
of fibers decreases to zero. (Figure 1A).

The spiking activity of a generic axon bundle was simulated by
stimulating the fibers with a background spontaneous firing rate of 100
spikes/second, upon which a pulse of increased activity was added. This
pulse of activity could relate to any kind of evoked activity in the
nervous system, such as a sensory stimulation, motor activity or a
spontaneous transient increase in population spiking activity. We chose
a gaussian pulse with an amplitude of 2900 spikes/second and a standard
deviation of 2.8ms.

To characterize the spatiotemporal dynamics of the evoked EFP, the
potential was calculated for several locations along the axon trunk.
Because we first want to study the effect of the rate pulse on the EFP,
we low-pass filtered the response the responses with a cut-off frequency
of 1kHz. Figures 1A-C all show these low-pass filtered responses. The
distribution of maximum amplitudes of these responses are shown by the
contour lines in Figure 1A. Surrounding the projection zone of the axon
bundle, they show the double-lobed shape typical of a dipole.

Shown alongside the contour lines are dots representing recording
locations, also colored according to the maximum amplitude at their
location. The waveforms recorded at these locations are shown in Figure
1B. The scale of the waveforms are indicated by the color of the trace.
The responses show a biphasic deflection that is elicited by the
population firing rate pulse. This deflection was visible at all
recording locations. It reverses polarity in the middle of the
bifurcation zone of the bundle (red box in Figure 1B). The polarity
reversal occurs by a decrease of the amplitude to zero and subsequent
reappearance with reversed polarity, as opposed to a polarity reversal
through a gradual shift in phase. This behaviour is also typical of a
dipolar field potential.

The point of the polarity reversal coincides roughly with the middle of
the projection zone. This means that the absolute value of the response
amplitude reaches a local minimum, just when the number of axonal fibers
reaches a maximum. To better understand how anatomical features and
response amplitude are related, we compared the value of the EFP
corresponding to the maximum absolute amplitude with the difference in
branchings and terminations (Figure 1C). The amplitude of the maximum
response (red trace) follows the distribution of branchings and
terminations(black histogram).

Along the nerve trunk the fiber density is constant, and the amplitude
is unchanged. As the axon bundle reaches its projection zone, the number
of bifurcations increases, and the response increases in amplitude. In
the middle of the projection zone, the number of bifurcations and
terminations cancels out. At the same depth, the amplitude crosses zero.
As the end of the projection zone approaches, the terminations outweigh
the bifurcations, and the amplitude increases, but the polarity is
reversed. This transition also corresponds to the traces in the red box
in Figure 1B. As the axon bundle ends, there are no longer any
bifurcations or terminations, and the amplitude decays toward zero.

@@

To understand the contributions related to individual spikes, we next
turned our attention to the high-frequency range. We filtered the EFP
responses with a high-pass filter, using the same cut-off frequency of
1kHz as before. The maximum amplitudes for this frequency band did not
show a double-lobe, but rather appeared like an ellipsoid centered at
the projection zone (Figure 1D). The waveforms are more disparate across
recording locations, suggesting a more local nature. There is no visible
polarity reversal (Figure 1E).

The high-frequency component changes its amplitude in the longitudinal
direction in accordance to the local fiber density (Figure 1F). It has
an approximately constant amplitude along the nerve trunk and then
increases in amplitude as the fiber density is increased by
bifurcations. As the fibers terminate and fiber density decreases, so
does the amplitude of the high-frequency component.

We have thus shown a qualitatively different behaviour in the low- and
high-frequency components of the EFP. Both appear to be related to
anatomy, though each in a different way. Together, these relationships
suggest that the particular branching and terminating structure of the
axon bundle gives rise to the dipolar field potential. Because dipolar
fields are associated with a longer spatial reach than the quadrupolar
fields typically associated with axonal spikes, we decided to
investigate the effect that axon structure has on the spatial reach of
the frequency components (Figure 2). To do this we performed a further
simulation of an axon bundle, this time omitting the bifurcations, and
letting the fibers terminate directly. In axial direction, the
low-frequency component has a further reach than the high frequency
component in the bifurcating case (Figure 2A), but not in the
non-bifurcating case (Figure 2C). In the radial direction, the situation
is different: the bifurcating (Figure 2B) and non-bifurcating (Figure
2D) case both show a faster decay of the high-frequency component than
of the low-frequency component. To assure that the effect was due to the
bifurcations themselves and not due to the fanned out structure of the
bifurcating bundle seen in Figures 2A and 2D, we here changed the layout
of both bundles to have constant widths.
